Sampling requirements to detect ecosystem change

### 2020-04-01 R/manuscript_draft.Rmd
made small changes to text throughout and added citations

---
bibliography: ChangepointSampling.bib
csl: ecology-letters.csl
editor_options:
  chunk_output_type: console
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount} \usepackage{soul}
  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \renewcommand{\thefootnote}{\fnsymbol{footnote}} \addtocounter{footnote}{-1} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = dirname(getwd()))
```

@@

@@

Environmental monitoring is a key component of managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage, which was reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are inefficient. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling, and where the objective is to detect ecosystem change.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection (citations). Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data. \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths. With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at. If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}

Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently. Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a,@McDonald-Madden2010a,@Lindenmayer2020a]. -->

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019d]. Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population. Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions [@Cohen2003]. Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a]. Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a,@Mallet2014]. In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed once per minute? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition at random locations along the core. Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018]. We hypothesize that random sampling will perform the worst at getting as close as possible to a "real" changepoint. Regular sampling is commonly used (citations) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g). In each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period. We describe each approach in more detail in the supplementary material and provide code.

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4). In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one. The distance to real changepoint served as performance diagnostic.

@@

We explored how each of these model parameters affected our ability to detect a change point. We simulate an entire time series to serve as the "true" data for comparison [@White2020]. We specifically study how the amount of samples included and the type of sampling affects the detection probability. For simulations, statistical power is the fraction of simulations that were able to detect a changepoint. We define an accurate changepont detection if an estimate is within five time points (over 100 time points) of the true changepoint. We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.

@@

source('R/sampling_regular.R')
source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

@@

# 2 - Calculate autocorrelation of real time serie
ar_vr <- acf(vr_time_series, lag.max = 5, plot = FALSE)$acf[2]

@@

In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size. Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time_plot}). For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.

@@

```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time}.\\label{fig:simulation_approaches_plot}'}

@@

p1  <- ggplot(myoutput_summary, aes(x=number_samples,y=mean_distance)) + geom_line()  + geom_ribbon(aes(ymin = mean_distance - 1.96*sd_distance, ymax = mean_distance + 1.96*sd_distance), fill = "grey80") +geom_line(size=1) + coord_cartesian(ylim = c(2.2, 50))  + 
  facet_wrap(~approach,
             labeller = label_glue('{.l}. {approach} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to real changepoint") #+
 # theme(legend.position = c(0.93, 0.72),
#       legend.background = element_rect(fill = "#ffffffaa", colour = NA),
#        legend.title=element_text(size=7), 
 #       legend.text=element_text(size=8)) +
#  guides(pch=guide_legend(title="Initial number\nof sample"))

@@

We tested different sampling methods on a real community time-series, from Lake Varese (IT). Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref). Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999,@Bruel2018]. Air temperature is now driving changes in plankton communities [@Bruel2018].

@@

We found that random sampling performed the worst, as changepoint analysis was left to chance. Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance). Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig. \ref{fig:case_study_number_sample} (c)). We show how iterative sampling changes the ordination, and the resulting changepoint, as more samples are added (Fig. \ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig. \ref{fig:case_study_number_sample} (c), square markers), the distance to the real changepoint was initially very high. Without any prior, the initial number sampled fell far away from the true changepoint.-->

@@

# Plot
p1 <- if(several_n2_init) {
  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method,
             labeller = label_glue('{.l}. {method} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.72),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
} else {
  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method,
             labeller = label_glue('{.l}. {method} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to real changepoint") 
}

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint. Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples. The method also work to detect further shifts (Fig. S4).

@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an interval approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

The approach is versatile: we converted the community data in a single vector of number of individuals per gram of sediment from the same sediment core (Fig. S5).  We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S6). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample} (c)). However, depending on the metric the user is interested in, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. \easton{I'm not sure what this means. What do you mean by it ``matches"?} \rosalie{is it clearer that way? I changed "match" to "corresponds" and added more details.} It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in observed individuals with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample. When sample analysis costs are high, then regular and iterative sampling are better than random sampling (Figs. \ref{fig:simulation_approaches_plot},S2,S3). However, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016]. However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in addressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.

>>>>

---
bibliography: [ChangepointSampling.bib, packages.bib]
csl: ecology-letters.csl
editor_options:
  chunk_output_type: console
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
  \usepackage[numbers]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount} \usepackage{soul}
  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite} \renewcommand{\thefootnote}{\fnsymbol{footnote}} \addtocounter{footnote}{-1} \newcommand\easton[1]{\textcolor{red}{#1}} \newcommand\rosalie[1]{\textcolor{blue}{#1}}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = dirname(getwd()))
# Optional code below to include package citations
# knitr::write_bib(c('dplyr', 'rmarkdown'), file = 'packages.bib')
# Use [@R-dplyr; @R-rmarkdown] in main text
```

@@

\linenumbers

@@

Environmental monitoring is a key component of understanding and managing ecosystems. Given that most monitoring efforts are still expensive and time-consuming, it is essential that monitoring programs are designed to be efficient and effective. In many situations, the expensive part of monitoring is not sample collection, but instead sample processing, which leads to only a subset of the samples being processed. For example, sediment or ice cores can be quickly obtained in the field, but they require weeks or months of processing in a laboratory setting. Standard sub-sampling approaches often involve equally-spaced sampling. We use simulations to show how many samples and which types of sampling approaches are the most effective in detecting ecosystem change. We test these ideas with a case study of Cladocera community assemblage, which was reconstructed from a sediment core. We demonstrate that standard approaches to sample processing are inefficient. For our case study, using an optimal sampling approach would have resulted in savings of 195 person-hours---thousands of dollars in labor costs. We also show that, compared with these standard approaches, fewer samples are typically needed to achieve high statistical power. We explain how our approach can be applied to monitoring programs that rely on video records, eDNA, remote sensing, and other common tools that allow re-sampling.

@@

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a; @White2019d; @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection [@Zhang2012]. Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a; @McDonald-Madden2010a; @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data. \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths. With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at. If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}

Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently. Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself [@Legg2006a;@McDonald-Madden2010a;@Lindenmayer2020a]. -->

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019d]. Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population. Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, sediment cores can be retrieved from lakes, allowing reconstruction of past ecological communities or conditions [@Cohen2003]. Similarly, environmental samples (e.g. water, soil) can be saved and processed later for composition, including eDNA [@Bohmann2014a]. Likewise, photo- or video-based monitoring can record snapshots of a system and be analyzed later [@OConnell2011a;@Mallet2014]. In each of these cases, decisions have to made about how much data to extract from the previously collected samples [@Zhang2012]. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed once per minute? As long as processing samples is expensive, these trade-offs will remain.

@@

The random sampling approach involves taking a set number of random points throughout the time series (Fig. \ref{fig:conceptual}c). In the context of sediment cores, this would mean analyzing community composition at random locations along the core. Random sampling is recommended in designs aimed at quantifying the average size of a population (spatial approach) [@Nad2018]. We hypothesize that random sampling will perform the worst in estimating the changepoint. Regular sampling is commonly used (e.g. @Bruel2018) and requires that a set number of samples be taken at regular intervals (Fig. \ref{fig:conceptual}d). Lastly, iterative sampling involves first taking a set number of samples and then iteratively adding samples until a pre-determined level of precision is achieved (Fig. \ref{fig:conceptual}e-g). For each scenario, we begin by sampling the first and last sample to ensure coverage of the whole time period. We describe each approach in more detail in the supplementary material and provide code.

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4). In order to test the performance, we detected the "true" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "true" one. The distance to true changepoint served as the performance diagnostic.

@@

We explored how each of these model parameters affected our ability to detect a change point. We simulate an entire time series to serve as the "true" data for comparison [@White2020]. We specifically study how the number of samples and the type of sampling affects the detection probability. For simulations, statistical power is the fraction of simulations that were able to detect a changepoint. We define an accurate changepont detection if an estimate is within five time points (over 100 time points) of the true changepoint. The minimum number of samples required is the number needed for 0.8 statistical power.

@@

source('R/sampling_regular.R')
source('R/sampling_random.R')
source('R/build_time_series.R')
#source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

@@

# 2 - Calculate autocorrelation of true time serie
ar_vr <- acf(vr_time_series, lag.max = 5, plot = FALSE)$acf[2]

@@

In line with theory on optimal monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size. Thus, the number of samples required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time_plot}). For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.

@@

```{r simulation_approaches_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint per total number of samples analyzed for simulations, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Each sample size and approach combination was simulated 50 times with the same parameters as Figure \\ref{fig:simulation_min_time_plot}. The error bars represent the middle 95% of the simlations.\\label{fig:simulation_approaches_plot}'}

@@

p1  <- ggplot(myoutput_summary, aes(x=number_samples,y=mean_distance)) + geom_line()  + geom_ribbon(aes(ymin = mean_distance - 1.96*sd_distance, ymax = mean_distance + 1.96*sd_distance), fill = "grey80") +geom_line(size=1) + coord_cartesian(ylim = c(2.2, 50))  + 
  facet_wrap(~approach,
             labeller = label_glue('{.l}. {approach} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to true changepoint") #+
 # theme(legend.position = c(0.93, 0.72),
#       legend.background = element_rect(fill = "#ffffffaa", colour = NA),
#        legend.title=element_text(size=7), 
 #       legend.text=element_text(size=8)) +
#  guides(pch=guide_legend(title="Initial number\nof sample"))

@@

We tested different sampling methods on a real community time-series, from Lake Varese (IT). Lake Varese is a small (14.8 km^2^), deep (z$_{max}$= 26m), monomictic lake, in the subalpine region of north-western Italy (238 m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed. Nutrient status was responsible for restructuration of the lake communities across trophic levels [@Crosta1999;@Bruel2018]. Air temperature is now driving changes in plankton communities [@Bruel2018].

@@

We found that random sampling performed the worst, as changepoint analysis was left to chance. Regular sampling provided good estimates from 8 samples (0-3 sample of absolute error), but detecting the true changepoint depended on the interval falling close to the true changepoint (i.e., left to chance). Iterative sampling, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint (Fig. \ref{fig:case_study_number_sample}c). We show how iterative sampling changes the ordination, and the resulting changepoint, as more samples are added (Fig. \ref{fig:case_study_iterative_sampling}) <!--Note that when the initial number of samples was set to 7 Fig. \ref{fig:case_study_number_sample} (c), square markers), the distance to the true changepoint was initially very high. Without any prior, the initial number sampled fell far away from the true changepoint.-->

@@

# Plot
p1 <- if(several_n2_init) {
  ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method,
             labeller = label_glue('{.l}. {method} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.72),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
} else {
  ggplot(myoutput[myoutput$target_cpt==1 & myoutput$n2_init == 5,], aes(final_n,diff_real)) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method,
             labeller = label_glue('{.l}. {method} sampling ')) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0)) +
  labs(x="Total number of samples analyzed", y="Distance to true changepoint") 
}

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be made as to which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at random points, random intervals, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint. Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}). The method also worked well to detect other changepoints of lower magnitude (Fig. S4). @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an interative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

As another application of our approach, we examined the same lake core data, but examined total abundance as opposed to community composition (Fig. S5). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S6). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the begining.

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d;@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012]. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occured, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample. When sample analysis costs are high, then regular and iterative sampling are better than random sampling (Figs. \ref{fig:simulation_approaches_plot},S2,S3). In addition, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015;@Doncaster2016]. However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implication in addressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.
